---
title: "Spatial Early Warning Signals Results Summary - SST and Heatwave Analysis"
date: "`r format(Sys.Date(), '%e %B %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
library(knitr)
plot.dir<- here::here("plots")

```

```{r plot-function,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE,fig.height = 10, fig.width = 8}
library(tidyverse)
library(plyr)
library(ecodata)
library(patchwork)
library(gridExtra)
library(earlywarnings)
library(plotly)

# read in all csvs
csv.dir <- here::here("data")
data.dir <- here::here("data//")

## SST prep 
sst_index<- ecodata::seasonal_oisst_anom %>%
  tidyr::separate(Var, c("Season", "a", "b", "c"), " ") %>% 
  dplyr::filter(!Time == 1982) %>% 
  dplyr::rename(anom.index = Value, 
         Year = Time) %>% 
  dplyr::select(-a, -b, -c, -Units)

### CCF
ccf_test<-function(epu_name, season_name, indicator, index){
  data.dir<-here::here("data//")
test<- index %>% 
  filter(EPU == epu_name, 
         Season == season_name)

sews <- read.csv(file = paste0(data.dir, paste0(epu_name,"_",season_name,"_",indicator,".csv"))) %>% 
  left_join(test) %>% 
  dplyr::mutate(anom.var = mean(variance) - variance, 
                anom.skew = mean(skewness) - skewness) %>% 
  tidyr::drop_na(anom.index)

ccf.values <- ccf(sews$anom.var, sews$anom.index, plot = FALSE)
ccf.df<-data.frame(lag=ccf.values$lag, CCF=ccf.values$acf) %>% 
  dplyr::mutate(val = abs(CCF)) %>% 
  dplyr::filter(val > 0.42) %>% 
  dplyr::select(!val) %>% 
  dplyr::mutate(comp = c("var~index"))


ccf.values2 <- ccf(sews$anom.skew, sews$anom.index, plot = FALSE)
ccf.df2<-data.frame(lag=ccf.values2$lag, CCF=ccf.values2$acf) %>% 
  dplyr::mutate(val = abs(CCF)) %>% 
   dplyr::filter(val > 0.42) %>% 
  select(!val)%>% 
  dplyr::mutate(comp = c("skew~index")) %>% 
  dplyr::bind_rows(ccf.df)

if ( nrow(ccf.df2) >0){
  ccf.df.grob <-  gridExtra::tableGrob(ccf.df2)
} else { 
  ccf.df.grob <-  gridExtra::tableGrob(list(c("No Lags detected using CCF")))
}

#ccf.df.grob <-  gridExtra::tableGrob(ccf.df2)
sews2<- sews %>% 
  tidyr::pivot_longer( cols = starts_with("anom"), 
                       names_to = "Variable", 
                       values_to = "Values") %>% 
  dplyr::mutate(Variable = factor(Variable, levels= c("anom.var", "anom.skew", "anom.index"),
                                  labels = c("Variance", "Skewness", "Index")))
p1<- sews2 %>%
  filter(EPU == epu_name,
         Season == season_name) %>%
  ggplot()+
  geom_line(aes(x = Year, y = Values))+
  geom_point(aes(x = Year, y = Values))+
  facet_wrap(.~Variable, ncol = 1, scales = "free")+
  ggtitle(paste(indicator, epu_name, season_name ))

gridExtra::grid.arrange(p1, ccf.df.grob,
             nrow=2,
             as.table=TRUE,
             heights=c(3,1))
}

```

# SST and Marine heatwaves {.tabset .tabset-fade}

## Monthly heatwave early warning signals

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(earlywarnings)
library(heatwaveR)
library(tidyverse)
library(readxl)
library(stringr)
library(readr)
library(lubridate)

raw.dir<- here::here("data-raw")

heatwave_gb_csv<-"GB_OISST - Vincent Saba - NOAA Federal.csv"
heatwave_gom_csv<-"GOM_OISST - Vincent Saba - NOAA Federal.csv"
heatwave_mab_csv<-"MAB_OISST - Vincent Saba - NOAA Federal.csv"

  gom<-read_csv(file.path(raw.dir,heatwave_gom_csv),
                 col_types = cols(temp = col_double(),t = col_date()))
  gb<-read_csv(file.path(raw.dir,heatwave_gb_csv),
                col_types = cols(temp = col_double(),t = col_date()))
  mab<-read_csv(file.path(raw.dir,heatwave_mab_csv),
                col_types = cols(temp = col_double(),t = col_date()))
  #GB
  ts <- heatwaveR::ts2clm(gb, climatologyPeriod = c("1982-01-01", "2010-12-31"))
  gb.mhw <- heatwaveR::detect_event(ts)
  gb.hw<- gb.mhw$event %>%
    dplyr::select(event_no, duration, date_start, date_peak, intensity_max, intensity_cumulative)%>%
    dplyr::mutate(EPU = "GB")
  #GOM
  ts <- heatwaveR::ts2clm(gom, climatologyPeriod = c("1982-01-01", "2010-12-31"))
  gom.mhw <- heatwaveR::detect_event(ts)
  gom.hw<- gom.mhw$event %>%
    dplyr::select(event_no, duration, date_start, date_peak, intensity_max, intensity_cumulative) %>%
    dplyr::mutate(EPU = "GOM")
  # MAB
  ts <- heatwaveR::ts2clm(mab, climatologyPeriod = c("1982-01-01", "2010-12-31"))
  mab.mhw <- heatwaveR::detect_event(ts)
  mab.hw<- mab.mhw$event %>%
    dplyr::select(event_no, duration, date_start, date_peak, intensity_max, intensity_cumulative) %>%
    dplyr::mutate(EPU = "MAB")
  # Cumulative intensity
  cum.intensity <- rbind(gb.hw, gom.hw, mab.hw) %>%
    dplyr::mutate(Mon = lubridate::month(date_peak), 
                  Year = lubridate::year(date_peak)) %>% 
    dplyr::group_by(EPU, Mon, Year) %>% 
    dplyr::summarise(Value = as.numeric(sum(intensity_cumulative)))
  pad.df <-as_tibble(expand.grid(Mon = 1:12, Year = 1982:2019, EPU = c("GOM", "GB", "MAB"))) %>% 
    dplyr::mutate(Value = c(0), 
                  Year = as.numeric(Year), 
                  Mon = as.numeric(Mon), 
                  EPU = as.character(EPU), 
                  Value = as.numeric(Value))%>% 
    dplyr::left_join( cum.intensity, by = c("Year", "Mon", "EPU"))
  cumulative <-  pad.df %>% 
    dplyr::mutate(Value.x = as.vector(Value.x),
                 Value.y = as.vector(Value.y)) %>% 
    dplyr::mutate(Value = dplyr::coalesce(pad.df$Value.y, pad.df$Value.x), 
                  Var = "cumulative intensity") %>% 
    dplyr::select(Year, Mon, Value, Var, EPU)
    
  #Max intensity
  max.intensity <- rbind(gb.hw, gom.hw, mab.hw) %>%
    dplyr::mutate(Mon = lubridate::month(date_peak), 
                  Year = lubridate::year(date_peak)) %>% 
    dplyr::rename(Value = intensity_max) %>%
    dplyr::mutate(Var = "maximum intensity")%>%
    group_by(Year, Mon, EPU) %>% 
    dplyr::filter(Value == max(Value))
 pad.df <-as_tibble(expand.grid(Mon = 1:12, Year = 1982:2019, EPU = c("GOM", "GB", "MAB"))) %>% 
    dplyr::mutate(Value = c(0), 
                  Year = as.numeric(Year), 
                  Mon = as.numeric(Mon), 
                  EPU = as.character(EPU), 
                  Value = as.numeric(Value))%>% 
    dplyr::left_join( max.intensity, by = c("Year", "Mon", "EPU"))
  maximum <-  pad.df %>% 
    dplyr::mutate(Value.x = as.vector(Value.x),
                 Value.y = as.vector(Value.y)) %>% 
    dplyr::mutate(Value = dplyr::coalesce(pad.df$Value.y, pad.df$Value.x), 
                  Var = "maximum intensity") %>% 
    dplyr::select(Year, Mon, Value, Var, EPU)

  heatwave<- rbind(cumulative, maximum) %>%
    dplyr::mutate(Time = make_date(Year, Mon)) %>% 
    dplyr:: mutate(Units = "degrees C")%>% 
    dplyr::filter(EPU == "GOM", 
                  Var == "cumulative intensity") 


p2 <- heatwave %>% 
  ggplot() +
  ggplot2::geom_point(aes(x=Time, y=Value))+
  ggplot2::geom_line(aes(x=Time, y=Value))+
  ggplot2::ggtitle("Monthly cumulative heatwaves 1982-2019")+
  ggplot2::ylab("Cumulative Intensity (degrees X days)")
p2


```


Time-series early warnings - analysis of monthly heatwave timeseries. Window size presented is 50% other window sizes were explored and can be found [here](https://kimberly-bastille.github.io/sp-warnings/summary_sst). This analysis uses the `generic_ews()` function in the `earlywarnings package`. 


```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}

heatwave<- heatwave %>% dplyr::mutate(Num = 1:dim(heatwave)[1]) %>% 
  dplyr::select(Time, Value)
ews <- earlywarnings::generic_ews(heatwave, winsize = 50, powerspectrum = TRUE)
trash<-dev.off()

#plot(ews)
ews.p <- ews %>%
  tidyr::pivot_longer(., cols = 2:9, names_to="ews.output") %>%
  dplyr::filter(ews.output == c("cv", "sk", "sd")) %>% 
   ggplot2::ggplot() +
   ggplot2::geom_line(aes(x=timeindex, y = value)) +
   ggplot2::xlim(4000,18000) +
   ggplot2::facet_wrap(~ews.output, ncol = 1, scales = "free")+
  ggplot2::ggtitle("Generic early warnings analysis of monthly heatwave index")
ews.p

```

### Summary 

From the monthly heatwave timeseries,it is clear that since ~2010 we have entered a period of more heatwaves that occur more often, last longer and are hotter than the past. The well documented 2012 heatwave in the Gulf of Maine can be seen in the Variance and Skewness calculated from the generic_ews() function in earlywarnings. 



#### Data Background

Heatwaves were calculated using the `heatwaveR` package and the flickering dataset produced shows the cumulative heatwave calculations for each month. If the heatwave occured in more than one month the peak temperature date (output from `heatwaveR`) was used. 



Data range = 1982 to 2019 and covers the Gulf of Maine. 





## Decomposition of the GOM SST spatial variance.

Spatial early warnings - Analysis or spatial variance for each monthly raster grid. 


```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
sews<- read.csv(file = paste0(data.dir, "GOM_sst.csv")) 

decompsews<- sews %>% 
  dplyr::select(variance) 

ts<- ts(decompsews, frequency=12)
decomp.ts<-stats::decompose(ts)

decomp.df<-   decomp.ts %>% 
  unlist(recursive = FALSE) %>% 
  enframe() %>% 
   separate(name, 
           into = c("name", "month"), 
           sep = "(?<=[A-Za-z])(?=[0-9])") %>% 
  drop_na(value) %>% 
  dplyr::filter(name != "type", 
                name != "figure") %>% 
  dplyr::mutate(value = as.numeric(value), 
               month = as.numeric(month), 
               name = fct_relevel(name, 
            "x", "trend", "seasonal", 
            "random"))

p1<-
  ggplot(decomp.df, aes(x = month,  y = value))+
  ggplot2::geom_line()+
  ecodata::geom_gls()+
  facet_wrap(name~., nrow = 4)
p1

```

### Summary 

Variance in sea surface temperature varies seasonally, meaning variance in the summer months is greater than the variance in the winter months. 

### Data Background 
SST data comes from NOAA Physical Science Laboratory [OISST V2 High Resolution Dataset](https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html). 

Data range = 1982 to 2019,  0.25 degree pixels masked to the EPUs of the NEUS shelf. This analysis only covers the Gulf of Maine EPU. 


Data gathered into monthly sst rasters using this [code](https://github.com/kimberly-bastille/sp-warnings/blob/master/R/get_monthly_sst.R)

