---
title: "Spatial Early Warning Signals Results Summary - SST example"
date: "`r format(Sys.Date(), '%e %B %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
library(knitr)
plot.dir<- here::here("plots")

```

```{r plot-function,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE,fig.height = 10, fig.width = 8}
library(tidyverse)
library(plyr)
library(ecodata)
library(patchwork)
library(gridExtra)
library(earlywarnings)
library(plotly)

# read in all csvs
csv.dir <- here::here("data")
data.dir <- here::here("data//")

## SST prep 
sst_index<- ecodata::seasonal_oisst_anom %>%
  tidyr::separate(Var, c("Season", "a", "b", "c"), " ") %>% 
  dplyr::filter(!Time == 1982) %>% 
  dplyr::rename(anom.index = Value, 
         Year = Time) %>% 
  dplyr::select(-a, -b, -c, -Units)

### CCF
ccf_test<-function(epu_name, season_name, indicator, index){
  data.dir<-here::here("data//")
test<- index %>% 
  filter(EPU == epu_name, 
         Season == season_name)

sews <- read.csv(file = paste0(data.dir, paste0(epu_name,"_",season_name,"_",indicator,".csv"))) %>% 
  left_join(test) %>% 
  dplyr::mutate(anom.var = mean(variance) - variance, 
                anom.skew = mean(skewness) - skewness) %>% 
  tidyr::drop_na(anom.index)

ccf.values <- ccf(sews$anom.var, sews$anom.index, plot = FALSE)
ccf.df<-data.frame(lag=ccf.values$lag, CCF=ccf.values$acf) %>% 
  dplyr::mutate(val = abs(CCF)) %>% 
  dplyr::filter(val > 0.42) %>% 
  dplyr::select(!val) %>% 
  dplyr::mutate(comp = c("var~index"))


ccf.values2 <- ccf(sews$anom.skew, sews$anom.index, plot = FALSE)
ccf.df2<-data.frame(lag=ccf.values2$lag, CCF=ccf.values2$acf) %>% 
  dplyr::mutate(val = abs(CCF)) %>% 
   dplyr::filter(val > 0.42) %>% 
  select(!val)%>% 
  dplyr::mutate(comp = c("skew~index")) %>% 
  dplyr::bind_rows(ccf.df)

if ( nrow(ccf.df2) >0){
  ccf.df.grob <-  gridExtra::tableGrob(ccf.df2)
} else { 
  ccf.df.grob <-  gridExtra::tableGrob(list(c("No Lags detected using CCF")))
}

#ccf.df.grob <-  gridExtra::tableGrob(ccf.df2)
sews2<- sews %>% 
  tidyr::pivot_longer( cols = starts_with("anom"), 
                       names_to = "Variable", 
                       values_to = "Values") %>% 
  dplyr::mutate(Variable = factor(Variable, levels= c("anom.var", "anom.skew", "anom.index"),
                                  labels = c("Variance", "Skewness", "Index")))
p1<- sews2 %>%
  filter(EPU == epu_name,
         Season == season_name) %>%
  ggplot()+
  geom_line(aes(x = Year, y = Values))+
  geom_point(aes(x = Year, y = Values))+
  facet_wrap(.~Variable, ncol = 1, scales = "free")+
  ggtitle(paste(indicator, epu_name, season_name ))

gridExtra::grid.arrange(p1, ccf.df.grob,
             nrow=2,
             as.table=TRUE,
             heights=c(3,1))
}

```

# SST and Marine heatwaves {.tabset .tabset-fade}

### Decomposition of the GOM SST spatial variance.
- monthly sst monthly variance

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
sews<- read.csv(file = paste0(data.dir, "GOM_sst.csv")) 

decompsews<- sews %>% 
  dplyr::select(variance) 

ts<- ts(decompsews, frequency=12)
decomp.ts<-stats::decompose(ts)

decomp.df<-   decomp.ts %>% 
  unlist(recursive = FALSE) %>% 
  enframe() %>% 
   separate(name, 
           into = c("name", "month"), 
           sep = "(?<=[A-Za-z])(?=[0-9])") %>% 
  drop_na(value) %>% 
  dplyr::filter(name != "type", 
                name != "figure") %>% 
  dplyr::mutate(value = as.numeric(value), 
               month = as.numeric(month), 
               name = fct_relevel(name, 
            "x", "trend", "seasonal", 
            "random"))

p1<-
  ggplot(decomp.df, aes(x = month,  y = value))+
  ggplot2::geom_line()+
  ecodata::geom_gls()+
  facet_wrap(name~., nrow = 4)
p1

```

## Annual heatwave ews 
### GOM Cumulative Marine Heatwave Index 
- Annual heatwave timeseries used in 2020 NE SOE 

```{r, echo=FALSE}
library(earlywarnings)

hw<- ecodata::heatwave %>% 
  filter(EPU == "GOM", 
         Var == "cumulative intensity") %>% 
  dplyr::select(Time, Value)

p2 <- hw %>% 
  ggplot() +
  ggplot2::geom_point(aes(x=Time, y=Value))+
  ggplot2::geom_line(aes(x=Time, y=Value))
p2


```


### GOM Cumulative Marine Heatwave Early Warnings 
earlywarnings::generic_ews() = Output includes ar1, sd, sk, kurt, cv, returnrate, densratio, acf1

<!-- ### window = 5% -->
<!-- - generic.ews() -->
<!-- - window size = 5  -->

<!-- ```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE} -->
<!-- ews <- earlywarnings::generic_ews(hw, winsize = 5) -->
<!-- trash<-dev.off() -->

<!-- #plot(ews) -->
<!-- ews.p <- ews %>% -->
<!--    tidyr::pivot_longer(., cols = 2:9, names_to="ews.output") %>% -->
<!--   dplyr::filter(ews.output == c("cv", "sk", "sd")) %>%  -->
<!--    ggplot2::ggplot() + -->
<!--    ggplot2::geom_line(aes(x=timeindex, y = value)) + -->
<!--    ggplot2::xlim(1982,2020) + -->
<!--    ggplot2::facet_wrap(~ews.output, ncol = 1, scales = "free") -->
<!-- ews.p -->

<!-- ``` -->


### window = 10%
- generic.ews()
- window size = 10 

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
ews <- earlywarnings::generic_ews(hw, winsize = 10)
trash<-dev.off()

#plot(ews)
ews.p <- ews %>%
   tidyr::pivot_longer(., cols = 2:9, names_to="ews.output") %>%
  dplyr::filter(ews.output == c("cv", "sk", "sd")) %>% 
   ggplot2::ggplot() +
   ggplot2::geom_line(aes(x=timeindex, y = value)) +
   ggplot2::xlim(1982,2020) +
   ggplot2::facet_wrap(~ews.output, ncol = 1, scales = "free")
ews.p

```




### window = 20%
- generic.ews()
- window size = 20 

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
ews <- earlywarnings::generic_ews(hw, winsize = 20)
trash<-dev.off()

#plot(ews)
ews.p <- ews %>%
   tidyr::pivot_longer(., cols = 2:9, names_to="ews.output") %>%
  dplyr::filter(ews.output == c("cv", "sk", "sd")) %>% 
   ggplot2::ggplot() +
   ggplot2::geom_line(aes(x=timeindex, y = value)) +
   ggplot2::xlim(1982,2020) +
   ggplot2::facet_wrap(~ews.output, ncol = 1, scales = "free")+
  ggplot2::xlab("Time")
ews.p

```


### window = 25%
- generic.ews()
- window size = 25 

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
ews <- earlywarnings::generic_ews(hw, winsize = 25)
trash<-dev.off()

#plot(ews)
ews.p <- ews %>%
   tidyr::pivot_longer(., cols = 2:9, names_to="ews.output") %>%
  dplyr::filter(ews.output == c("cv", "sk", "sd")) %>% 
   ggplot2::ggplot() +
   ggplot2::geom_line(aes(x=timeindex, y = value)) +
   ggplot2::xlim(1982,2020) +
   ggplot2::facet_wrap(~ews.output, ncol = 1, scales = "free")
ews.p

```

### window = 30%
- generic.ews()
- window size = 30 

```{r, echo=FALSE,eval=TRUE, message=FALSE, warning=FALSE}
ews <- earlywarnings::generic_ews(hw, winsize = 30)
trash<-dev.off()

#plot(ews)
ews.p <- ews %>%
   tidyr::pivot_longer(., cols = 2:9, names_to="ews.output") %>%
  dplyr::filter(ews.output == c("cv", "sk", "sd")) %>% 
   ggplot2::ggplot() +
   ggplot2::geom_line(aes(x=timeindex, y = value)) +
   ggplot2::xlim(1982,2020) +
   ggplot2::facet_wrap(~ews.output, ncol = 1, scales = "free")
ews.p

```

## Monthly heatwave ews

```{r, echo=FALSE}
library(earlywarnings)
library(heatwaveR)
library(tidyverse)
library(readxl)
library(stringr)
library(readr)
library(lubridate)

raw.dir<- here::here("data-raw")

heatwave_gb_csv<-"GB_OISST - Vincent Saba - NOAA Federal.csv"
heatwave_gom_csv<-"GOM_OISST - Vincent Saba - NOAA Federal.csv"
heatwave_mab_csv<-"MAB_OISST - Vincent Saba - NOAA Federal.csv"

  gom<-read_csv(file.path(raw.dir,heatwave_gom_csv),
                 col_types = cols(temp = col_double(),t = col_date()))
  gb<-read_csv(file.path(raw.dir,heatwave_gb_csv),
                col_types = cols(temp = col_double(),t = col_date()))
  mab<-read_csv(file.path(raw.dir,heatwave_mab_csv),
                col_types = cols(temp = col_double(),t = col_date()))
  #GB
  ts <- heatwaveR::ts2clm(gb, climatologyPeriod = c("1982-01-01", "2010-12-31"))
  gb.mhw <- heatwaveR::detect_event(ts)
  gb.hw<- gb.mhw$event %>%
    dplyr::select(event_no, duration, date_start, date_peak, intensity_max, intensity_cumulative)%>%
    dplyr::mutate(EPU = "GB")
  #GOM
  ts <- heatwaveR::ts2clm(gom, climatologyPeriod = c("1982-01-01", "2010-12-31"))
  gom.mhw <- heatwaveR::detect_event(ts)
  gom.hw<- gom.mhw$event %>%
    dplyr::select(event_no, duration, date_start, date_peak, intensity_max, intensity_cumulative) %>%
    dplyr::mutate(EPU = "GOM")
  # MAB
  ts <- heatwaveR::ts2clm(mab, climatologyPeriod = c("1982-01-01", "2010-12-31"))
  mab.mhw <- heatwaveR::detect_event(ts)
  mab.hw<- mab.mhw$event %>%
    dplyr::select(event_no, duration, date_start, date_peak, intensity_max, intensity_cumulative) %>%
    dplyr::mutate(EPU = "MAB")
  # Cumulative intensity
  cum.intensity <- rbind(gb.hw, gom.hw, mab.hw) %>%
    dplyr::mutate(Mon = lubridate::month(date_peak), 
                  Year = lubridate::year(date_peak)) %>% 
    dplyr::group_by(EPU, Mon, Year) %>% 
    dplyr::summarise(Value = as.numeric(sum(intensity_cumulative)))
  pad.df <-as_tibble(expand.grid(Mon = 1:12, Year = 1982:2019, EPU = c("GOM", "GB", "MAB"))) %>% 
    dplyr::mutate(Value = c(0), 
                  Year = as.numeric(Year), 
                  Mon = as.numeric(Mon), 
                  EPU = as.character(EPU), 
                  Value = as.numeric(Value))%>% 
    dplyr::left_join( cum.intensity, by = c("Year", "Mon", "EPU"))
  cumulative <-  pad.df %>% 
    dplyr::mutate(Value.x = as.vector(Value.x),
                 Value.y = as.vector(Value.y)) %>% 
    dplyr::mutate(Value = dplyr::coalesce(pad.df$Value.y, pad.df$Value.x), 
                  Var = "cumulative intensity") %>% 
    dplyr::select(Year, Mon, Value, Var, EPU)
    
  #Max intensity
  max.intensity <- rbind(gb.hw, gom.hw, mab.hw) %>%
    dplyr::mutate(Mon = lubridate::month(date_peak), 
                  Year = lubridate::year(date_peak)) %>% 
    dplyr::rename(Value = intensity_max) %>%
    dplyr::mutate(Var = "maximum intensity")%>%
    group_by(Year, Mon, EPU) %>% 
    dplyr::filter(Value == max(Value))
 pad.df <-as_tibble(expand.grid(Mon = 1:12, Year = 1982:2019, EPU = c("GOM", "GB", "MAB"))) %>% 
    dplyr::mutate(Value = c(0), 
                  Year = as.numeric(Year), 
                  Mon = as.numeric(Mon), 
                  EPU = as.character(EPU), 
                  Value = as.numeric(Value))%>% 
    dplyr::left_join( max.intensity, by = c("Year", "Mon", "EPU"))
  maximum <-  pad.df %>% 
    dplyr::mutate(Value.x = as.vector(Value.x),
                 Value.y = as.vector(Value.y)) %>% 
    dplyr::mutate(Value = dplyr::coalesce(pad.df$Value.y, pad.df$Value.x), 
                  Var = "maximum intensity") %>% 
    dplyr::select(Year, Mon, Value, Var, EPU)

  heatwave<- rbind(cumulative, maximum) %>%
    dplyr::mutate(Time = make_date(Year, Mon)) %>% 
    dplyr:: mutate(Units = "degrees C")%>% 
    dplyr::filter(EPU == "GOM", 
                  Var == "cumulative intensity") 


p2 <- heatwave %>% 
  ggplot() +
  ggplot2::geom_point(aes(x=Time, y=Value))+
  ggplot2::geom_line(aes(x=Time, y=Value))
p2


```


### window = 15%
- generic.ews()
- window size = 15 

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}

heatwave<- heatwave %>% dplyr::mutate(Num = 1:dim(heatwave)[1]) %>% 
  dplyr::select(Time, Value)
ews <- earlywarnings::generic_ews(heatwave, winsize = 15)
trash<-dev.off()

#plot(ews)
ews.p <- ews %>%
  tidyr::pivot_longer(., cols = 2:9, names_to="ews.output") %>%
  dplyr::filter(ews.output == c("cv", "sk", "sd")) %>% 
   ggplot2::ggplot() +
   ggplot2::geom_line(aes(x=timeindex, y = value)) +
   ggplot2::xlim(6000,18000) +
   ggplot2::facet_wrap(~ews.output, ncol = 1, scales = "free")
ews.p

```




### window = 25%
- generic.ews()
- window size = 25 

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
ews <- earlywarnings::generic_ews(heatwave, winsize = 25)
trash<-dev.off()

#plot(ews)
ews.p <- ews %>%
   tidyr::pivot_longer(., cols = 2:9, names_to="ews.output") %>%
  dplyr::filter(ews.output == c("cv", "sk", "sd")) %>% 
   ggplot2::ggplot() +
   ggplot2::geom_line(aes(x=timeindex, y = value)) +
   ggplot2::xlim(6000,18000) +
   ggplot2::facet_wrap(~ews.output, ncol = 1, scales = "free")+
  ggplot2::xlab("Time")
ews.p

```



### window = 50%
- generic.ews()
- window size = 50 

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
ews <- earlywarnings::generic_ews(heatwave, winsize = 50)
trash<-dev.off()

#plot(ews)
ews.p <- ews %>%
   tidyr::pivot_longer(., cols = 2:9, names_to="ews.output") %>%
  dplyr::filter(ews.output == c("cv", "sk", "sd")) %>% 
   ggplot2::ggplot() +
   ggplot2::geom_line(aes(x=timeindex, y = value)) +
   ggplot2::xlim(6000,18000) +
   ggplot2::facet_wrap(~ews.output, ncol = 1, scales = "free")+
  ggplot2::xlab("Time")
ews.p

```