---
title: "Spatial Early Warning Signals Results Summary - SST and Heatwave Analysis"
date: "`r format(Sys.Date(), '%e %B %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
library(knitr)
plot.dir<- here::here("plots")

```

```{r plot-function,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE,fig.height = 10, fig.width = 8}
library(tidyverse)
library(plyr)
library(ecodata)
library(patchwork)
library(gridExtra)
library(earlywarnings)
library(plotly)
library(heatwaveR)
library(readxl)
library(readr)

# read in all csvs
csv.dir <- here::here("data")
data.dir <- here::here("data//")

raw.dir<- here::here("data-raw")

```

# SST and Marine heatwaves {.tabset .tabset-fade}

### Summary

Question 1: Is there an early warning signal leading up to the 2012 heatwave event? 
Answer 1: I don't think so. Based on Figure 2. It looks like the change in variance and skewness occur during/after the events with a smaller reaction to a 2006 event occuring earlier on in the ts. So I don't think that this is showing up what we are looking for. 

Question 2: How can we link the heatwaves EWS to SST SEWS?
Answer 2: Both show seasonality in variation - Greater variation in the summer months and lower variation in the winter months. The SD is increasing in heatwaves EWS but relative variation (CV) is not. The Spatial Early Warnings analysis calculates variance using the base `r` function `var`. 

## Monthly heatwave early warning signals

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}

heatwave_gb_csv<-"GB_OISST - Vincent Saba - NOAA Federal.csv"
heatwave_gom_csv<-"GOM_OISST - Vincent Saba - NOAA Federal.csv"
heatwave_mab_csv<-"MAB_OISST - Vincent Saba - NOAA Federal.csv"

  gom<-read_csv(file.path(raw.dir,heatwave_gom_csv),
                 col_types = cols(temp = col_double(),t = col_date()))
  gb<-read_csv(file.path(raw.dir,heatwave_gb_csv),
                col_types = cols(temp = col_double(),t = col_date()))
  mab<-read_csv(file.path(raw.dir,heatwave_mab_csv),
                col_types = cols(temp = col_double(),t = col_date()))
  #GB
  ts <- heatwaveR::ts2clm(gb, climatologyPeriod = c("1982-01-01", "2010-12-31"))
  gb.mhw <- heatwaveR::detect_event(ts)
  gb.hw<- gb.mhw$event %>%
    dplyr::select(event_no, duration, date_start, date_peak, intensity_max, intensity_cumulative)%>%
    dplyr::mutate(EPU = "GB")
  #GOM
  ts <- heatwaveR::ts2clm(gom, climatologyPeriod = c("1982-01-01", "2010-12-31"))
  gom.mhw <- heatwaveR::detect_event(ts)
  gom.hw<- gom.mhw$event %>%
    dplyr::select(event_no, duration, date_start, date_peak, intensity_max, intensity_cumulative) %>%
    dplyr::mutate(EPU = "GOM")
  # MAB
  ts <- heatwaveR::ts2clm(mab, climatologyPeriod = c("1982-01-01", "2010-12-31"))
  mab.mhw <- heatwaveR::detect_event(ts)
  mab.hw<- mab.mhw$event %>%
    dplyr::select(event_no, duration, date_start, date_peak, intensity_max, intensity_cumulative) %>%
    dplyr::mutate(EPU = "MAB")
  # Cumulative intensity
  cum.intensity <- rbind(gb.hw, gom.hw, mab.hw) %>%
    dplyr::mutate(Mon = lubridate::month(date_peak), 
                  Year = lubridate::year(date_peak)) %>% 
    dplyr::group_by(EPU, Mon, Year) %>% 
    dplyr::summarise(Value = as.numeric(sum(intensity_cumulative)))
  pad.df <-as_tibble(expand.grid(Mon = 1:12, Year = 1982:2019, EPU = c("GOM", "GB", "MAB"))) %>% 
    dplyr::mutate(Value = c(0), 
                  Year = as.numeric(Year), 
                  Mon = as.numeric(Mon), 
                  EPU = as.character(EPU), 
                  Value = as.numeric(Value))%>% 
    dplyr::left_join( cum.intensity, by = c("Year", "Mon", "EPU"))
  cumulative <-  pad.df %>% 
    dplyr::mutate(Value.x = as.vector(Value.x),
                 Value.y = as.vector(Value.y)) %>% 
    dplyr::mutate(Value = dplyr::coalesce(pad.df$Value.y, pad.df$Value.x), 
                  Var = "cumulative intensity") %>% 
    dplyr::select(Year, Mon, Value, Var, EPU)
    
  #Max intensity
  max.intensity <- rbind(gb.hw, gom.hw, mab.hw) %>%
    dplyr::mutate(Mon = lubridate::month(date_peak), 
                  Year = lubridate::year(date_peak)) %>% 
    dplyr::rename(Value = intensity_max) %>%
    dplyr::mutate(Var = "maximum intensity")%>%
    group_by(Year, Mon, EPU) %>% 
    dplyr::filter(Value == max(Value))
 pad.df <-as_tibble(expand.grid(Mon = 1:12, Year = 1982:2019, EPU = c("GOM", "GB", "MAB"))) %>% 
    dplyr::mutate(Value = c(0), 
                  Year = as.numeric(Year), 
                  Mon = as.numeric(Mon), 
                  EPU = as.character(EPU), 
                  Value = as.numeric(Value))%>% 
    dplyr::left_join( max.intensity, by = c("Year", "Mon", "EPU"))
  maximum <-  pad.df %>% 
    dplyr::mutate(Value.x = as.vector(Value.x),
                 Value.y = as.vector(Value.y)) %>% 
    dplyr::mutate(Value = dplyr::coalesce(pad.df$Value.y, pad.df$Value.x), 
                  Var = "maximum intensity") %>% 
    dplyr::select(Year, Mon, Value, Var, EPU)

  heatwave<- rbind(cumulative, maximum) %>%
    dplyr::mutate(Time = lubridate::make_date(Year, Mon)) %>% 
    dplyr::mutate(Units = "degrees C")%>% 
    dplyr::filter(EPU == "GOM", 
                  Var == "cumulative intensity") 


p2 <- heatwave %>% 
  ggplot() +
  ggplot2::geom_point(aes(x=Time, y=Value))+
  ggplot2::geom_line(aes(x=Time, y=Value))+
  ecodata::geom_gls(aes(x=Time, y=Value))+
  ggplot2::ggtitle("Monthly cumulative heatwaves 1982-2019")+
  ggplot2::ylab("Cumulative Intensity (degrees X days)")
p2


```


Time-series early warnings - analysis of monthly heatwave timeseries. Window size presented is 50% other window sizes were explored and can be found [here](https://kimberly-bastille.github.io/sp-warnings/summary_sst). This analysis uses the `generic_ews()` function in the `earlywarnings package`. 


**RED LINE = 2012 MAJOR HEATWAVE EVENT**
```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}

heatwave<- heatwave %>% dplyr::mutate(Num = 1:dim(heatwave)[1]) %>% 
  dplyr::select(Num, Value)
ews <- earlywarnings::generic_ews(heatwave, winsize = 50, powerspectrum = TRUE)
trash<-dev.off()

#plot(ews)
ews.p <- ews %>%
  tidyr::pivot_longer(., cols = 2:9, names_to="ews.output") %>%
  dplyr::filter(ews.output == c("cv", "sk", "sd")) %>% 
   ggplot2::ggplot() +
   ggplot2::geom_line(aes(x=timeindex, y = value)) +
   ggplot2::geom_vline(xintercept = 365, color = "red")+
   #ggplot2::xlim(4000,18000) +
   ggplot2::facet_wrap(~ews.output, ncol = 1, scales = "free_y")+
   ggplot2::ggtitle("Generic early warnings analysis of monthly heatwave index")
ews.p

```


```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}

decompews<- ews %>% 
  dplyr::select(sd) 

# decompheat <- heatwave %>% 
#   dplyr::select(Value) 

ts<- ts(decompews, frequency=12)
decomp.ts<-stats::decompose(ts)

decomp.df<-   decomp.ts %>% 
  unlist(recursive = FALSE) %>% 
  enframe() %>% 
   separate(name, 
           into = c("name", "month"), 
           sep = "(?<=[A-Za-z])(?=[0-9])") %>% 
  drop_na(value) %>% 
  dplyr::filter(name != "type", 
                name != "figure") %>% 
  dplyr::mutate(value = as.numeric(value), 
               month = (as.numeric(month)+227), 
               name = fct_relevel(name, 
            "x", "trend", "seasonal", 
            "random"))

p1<-
  ggplot(decomp.df, aes(x = month,  y = value))+
  ggplot2::geom_line()+
  ecodata::geom_gls()+
  facet_wrap(name~., nrow = 4, scales = "free_y")+
  ggplot2::geom_vline(xintercept = 365, color = "red")+
  ggplot2::ggtitle("Decomposed EWS Standard Deviation")
p1

```


```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}

decompews<- ews %>% 
  dplyr::select(cv) 

# decompheat <- heatwave %>% 
#   dplyr::select(Value) 

ts<- ts(decompews, frequency=12)
decomp.ts<-stats::decompose(ts)

decomp.df<-   decomp.ts %>% 
  unlist(recursive = FALSE) %>% 
  enframe() %>% 
   separate(name, 
           into = c("name", "month"), 
           sep = "(?<=[A-Za-z])(?=[0-9])") %>% 
  drop_na(value) %>% 
  dplyr::filter(name != "type", 
                name != "figure") %>% 
  dplyr::mutate(value = as.numeric(value), 
               month = (as.numeric(month)+227), 
               name = fct_relevel(name, 
            "x", "trend", "seasonal", 
            "random"))

p1<-
  ggplot(decomp.df, aes(x = month,  y = value))+
  ggplot2::geom_line()+
  ecodata::geom_gls()+
  facet_wrap(name~., nrow = 4, scales = "free")+
  ggplot2::geom_vline(xintercept = 365, color = "red")+
  ggplot2::ggtitle("Decomposed EWS CV")
p1

```




#### Data Background

Heatwaves were calculated using the `heatwaveR` package and the flickering dataset produced shows the cumulative heatwave calculations for each month. If the heatwave occured in more than one month the peak temperature date (output from `heatwaveR`) was used. 

Data range = 1982 to 2019 and covers the Gulf of Maine. 


## Decomposition of the GOM SST spatial variance.

Spatial early warnings - Analysis or spatial variance for each monthly raster grid. 


```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
sews<- read.csv(file = paste0(data.dir, "GOM_sst.csv")) 

decompsews<- sews %>% 
  dplyr::select(variance) 

ts<- ts(decompsews, frequency=12)
decomp.ts<-stats::decompose(ts)

decomp.df<-   decomp.ts %>% 
  unlist(recursive = FALSE) %>% 
  enframe() %>% 
   separate(name, 
           into = c("name", "month"), 
           sep = "(?<=[A-Za-z])(?=[0-9])") %>% 
  drop_na(value) %>% 
  dplyr::filter(name != "type", 
                name != "figure") %>% 
  dplyr::mutate(value = as.numeric(value), 
               month = as.numeric(month), 
               name = fct_relevel(name, 
            "x", "trend", "seasonal", 
            "random"))

p1<-
  ggplot(decomp.df, aes(x = month,  y = value))+
  ggplot2::geom_line()+
  ecodata::geom_gls()+
  facet_wrap(name~., nrow = 4)
p1

```


### Data Background 
SST data comes from NOAA Physical Science Laboratory [OISST V2 High Resolution Dataset](https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html). 

Data range = 1982 to 2019,  0.25 degree pixels masked to the EPUs of the NEUS shelf. This analysis only covers the Gulf of Maine EPU. 


Data gathered into monthly sst rasters using this [code](https://github.com/kimberly-bastille/sp-warnings/blob/master/R/get_monthly_sst.R)

