
---
title: "Spatial Early Warning Signals Results Summary"
date: "`r format(Sys.Date(), '%e %B %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
plot.dir<- here::here("plots")

```


```{r plot-function,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(plyr)
library(ecodata)
library(patchwork)
library(gridExtra)


# read in all csvs
csv.dir <- here::here("data")
data.dir <- here::here("data//")
# epu_name = "MAB"
# season_name = "spring"
# indicator = "pp"


plot30plus<- function(epu_name, season_name, indicator){
  
  fname<- list.files(csv.dir, pattern = paste0(indicator,'.csv'),full.names=TRUE)
  dat <- ldply(fname, read.csv)
  epu.season<-dat %>% 
    filter(EPU == epu_name) %>% 
    filter(season == season_name)
  p1<-epu.season %>%   
    ggplot(aes(x = Year, y = variance))+
    geom_point(aes(x = Year, y = variance, color = "var"))+
    geom_line(aes(x = Year, y = variance, color = "var"))+
    geom_point(aes(x = Year, y = skewness, color = "skew"))+
    geom_line(aes(x = Year, y = skewness, color = "skew"))+
    ecodata::geom_gls(aes(x = Year, y = variance))+
    ecodata::geom_gls(aes(x = Year, y = skewness))+
    theme_ts()+
    theme_facet()+
    theme(axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(), 
          axis.title.x = element_blank())+
    ggtitle(paste(epu_name, season_name, indicator, " SEWS"))
  p1
}

plot30less<- function(epu_name, season_name, indicator){
  
  fname<- list.files(csv.dir, pattern = paste0(indicator,'.csv'),full.names=TRUE)
  dat <- ldply(fname, read.csv)
  epu.season<-dat %>% 
    filter(EPU == epu_name) %>% 
    filter(season == season_name)
  p1<-epu.season %>%   
    ggplot()+
    geom_point(aes(x = Year, y = variance, color = "var"))+
    geom_line(aes(x = Year, y = variance, color = "var"))+
    geom_point(aes(x = Year, y = skewness, color = "skew"))+
    geom_line(aes(x = Year, y = skewness, color = "skew"))+
    ecodata::geom_gls(aes(x = Year, y = variance), linetype = "dashed")+
    ecodata::geom_gls(aes(x = Year, y = skewness), linetype = "dashed")+
    theme_ts()+
    theme(axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(), 
          axis.title.x = element_blank())+
    theme_facet()+
    ggtitle(paste(epu_name, season_name, indicator, " SEWS"))
  p1
}

## CCF for PPD and ZOO
pp_zoo_test<-function(epu_name, season_name, indicator){
zoo <- read.csv(here::here("data/zooplankton.csv")) %>% 
    filter(EPU == epu_name,
           season == season_name,
           zoo_name == "Large") # calanus

pp <- read.csv(file = paste0(data.dir, paste0(epu_name,"_",season_name,"_",indicator,".csv"))) 

intsect<-intersect(zoo$year, pp$Year)

pp1<- pp %>% filter(Year %in% intsect)
zoo1<-zoo %>% filter(year %in% intsect)

test1<-cor.test(pp1$skewness,  zoo1$anom1, method = "pearson")
test2<-cor.test(pp1$variance,  zoo1$anom1, method = "pearson")

p1<- zoo1 %>%
  ggplot()+
  geom_line(aes(x =  year, y = anom1, color = "Zoo abundance"))+
  geom_line(aes(x= pp1$Year, y = pp1$variance, color = "PP Variance"))+
  geom_line(aes(x= pp1$Year, y = pp1$skewness, color = "PP Skewness"))+
  ecodata::geom_gls(aes(x = pp1$Year, y = pp1$variance), linetype = "dashed")+
  ecodata::geom_gls(aes(x = pp1$Year, y = pp1$skewness), linetype = "dashed")+
  ggtitle(paste("PPD~Zoo CCF", epu_name, season_name ))
p1
#test1,test2
}

## PPD prep
pp <- ecodata::chl_pp %>% 
  filter(str_detect(Var, "MONTHLY_PPD"), 
         !str_detect(Time, "1997")) %>% 
  tidyr::extract(Time, into = c("Year", "Month"), "(.{4})(.{2})") %>% 
  mutate(
    Season = case_when(
      Month %in% c("10", "11", "12") ~ "fall",
      Month %in% c("01", "02", "03")  ~ "winter",
      Month %in% c("04", "05", "06")  ~ "spring",
      TRUE ~ "summer"), 
    Year = as.numeric(Year))%>% 
  dplyr::group_by(Season, EPU, Year) %>% 
  dplyr::summarise(index = mean(Value)) %>% 
  dplyr::mutate(anom.index = mean(index) - index) %>% 
  ungroup()

### CCF
ccf_test<-function(epu_name, season_name, indicator){
  data.dir<-here::here("data//")
test<- pp %>% 
  filter(EPU == epu_name, 
         Season == season_name)

sews <- read.csv(file = paste0(data.dir, paste0(epu_name,"_",season_name,"_",indicator,".csv"))) %>% 
  left_join(test) %>% 
  dplyr::mutate(anom.var = mean(variance) - variance, 
                anom.skew = mean(skewness) - skewness) 

ccf.values <- ccf(sews$anom.var, sews$anom.index, plot = FALSE)
ccf.df<-data.frame(lag=ccf.values$lag, CCF=ccf.values$acf) %>% 
  dplyr::mutate(val = abs(CCF)) %>% 
  dplyr::filter(val >= 0.42) %>% 
  dplyr::select(!val) %>% 
  dplyr::mutate(comp = c("var~index"))


ccf.values2 <- ccf(sews$anom.skew, sews$anom.index, plot = FALSE)
ccf.df2<-data.frame(lag=ccf.values2$lag, CCF=ccf.values2$acf) %>% 
  dplyr::mutate(val = abs(CCF)) %>% 
  filter(val >= 0.42) %>% 
  select(!val)%>% 
  dplyr::mutate(comp = c("skew~index")) %>% 
  dplyr::bind_rows(ccf.df)

ccf.df.grob <-  gridExtra::tableGrob(ccf.df2)
sews2<- sews %>% 
  tidyr::pivot_longer( cols = starts_with("anom"), 
                       names_to = "Variable", 
                       values_to = "Values") %>% 
  dplyr::mutate(Variable = factor(Variable, levels= c("anom.var", "anom.skew", "anom.index"),
                                  labels = c("Variance", "Skewness", "Index")))
p1<- sews2 %>%
  filter(EPU == epu_name,
         Season == season_name) %>%
  ggplot()+
  geom_line(aes(x = Year, y = Values))+
  geom_point(aes(x = Year, y = Values))+
  facet_wrap(.~Variable, ncol = 1, scales = "free")+
  ggtitle(paste(indicator, epu_name, season_name ))

grid.arrange(p1, ccf.df.grob,
             nrow=2,
             as.table=TRUE,
             heights=c(3,1))
}

```




# Variance and skewness by epu and indicator {.tabset .tabset-fade}

## SST {.tabset .tabset-fade}

### MAB
```{r mab-sst, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30plus("MAB", "winter", "sst")
p2<-plot30plus("MAB", "spring", "sst")
p3<-plot30plus("MAB", "summer", "sst")
p4<-plot30plus("MAB", "fall", "sst")

(p1 | p2)/(p3 | p4)
```

### GOM
```{r GOM-sst, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30plus("GOM", "winter", "sst")
p2<-plot30plus("GOM", "spring", "sst")
p3<-plot30plus("GOM", "summer", "sst")
p4<-plot30plus("GOM", "fall", "sst")

(p1 | p2)/(p3 | p4)
```

### GB
```{r GB-sst, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE,fig.width=11, fig.height=12} 
p1<-plot30plus("GB", "winter", "sst")
p2<-plot30plus("GB", "spring", "sst")
p3<-plot30plus("GB", "summer", "sst")
p4<-plot30plus("GB", "fall", "sst")

(p1 | p2)/(p3 | p4)
```


### SST Trends in variance and skewness

```{r trends,echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
fname<- list.files(csv.dir, pattern = 'sst.csv',full.names=TRUE)
  sst <- ldply(fname, read.csv)
  sst %>% 
  ggplot(aes(x = Year, y = variance))+
  geom_point()+
  geom_line()+
  facet_grid(EPU~season, scales = "free")+
  geom_gls()+
  theme_ts()+
  theme_facet()+
  ggtitle("OISST variance")
  
  sst %>% 
  ggplot(aes(x = Year, y = skewness))+
  geom_point()+
  geom_line()+
  facet_grid(EPU~season, scales = "free")+
  geom_gls()+
  theme_ts()+
  theme_facet()+
  ggtitle("OISST skewness")
  
```


## CHL a {.tabset .tabset-fade}

### MAB
```{r mab-chl, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30less("MAB", "winter", "chl")
p2<-plot30less("MAB", "spring", "chl")
p3<-plot30less("MAB", "summer", "chl")
p4<-plot30less("MAB", "fall", "chl")

(p1 | p2)/(p3 | p4)
```

### GOM
```{r GOM-chl, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30less("GOM", "winter", "chl")
p2<-plot30less("GOM", "spring", "chl")
p3<-plot30less("GOM", "summer", "chl")
p4<-plot30less("GOM", "fall", "chl")

(p1 | p2)/(p3 | p4)
```

### GB
```{r GB-chl, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30less("GB", "winter", "chl")
p2<-plot30less("GB", "spring", "chl")
p3<-plot30less("GB", "summer", "chl")
p4<-plot30less("GB", "fall", "chl")

(p1 | p2)/(p3 | p4)
```


## PPD {.tabset .tabset-fade}

### CCF 
```{r mab-pp-ccf, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
##### What does this mean??
ccf_test("MAB", "winter", "pp") 
ccf_test("MAB", "spring", "pp") 
ccf_test("MAB", "summer", "pp") 
ccf_test("MAB", "fall", "pp") 
```

```{r gb-pp-ccf, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
ccf_test("GB", "winter", "pp") 
ccf_test("GB", "spring", "pp") 
ccf_test("GB", "summer", "pp") 
ccf_test("GB", "fall", "pp") 
```

```{r gom-pp-ccf, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
ccf_test("GOM", "winter", "pp") 
ccf_test("GOM", "spring", "pp") 
ccf_test("GOM", "summer", "pp") 
ccf_test("GOM", "fall", "pp") 
```





### MAB

#### SEWS
```{r mab-pp, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30less("MAB", "winter", "pp")
p2<-plot30less("MAB", "spring", "pp")
p3<-plot30less("MAB", "summer", "pp")
p4<-plot30less("MAB", "fall", "pp")

(p1 | p2)/(p3 | p4)
```

#### PP ~ Zoo abundance
```{r mab-pp-zoo, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-pp_zoo_test("MAB", "winter", "pp")
p2<-pp_zoo_test("MAB", "spring", "pp")
p3<-pp_zoo_test("MAB", "summer", "pp")
p4<-pp_zoo_test("MAB", "fall", "pp")

(p1 | p2)/(p3 | p4)
```


### GOM
```{r GOM-pp, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30less("GOM", "winter", "pp")
p2<-plot30less("GOM", "spring", "pp")
p3<-plot30less("GOM", "summer", "pp")
p4<-plot30less("GOM", "fall", "pp")

(p1 | p2)/(p3 | p4)
```

#### PP ~ Zoo abundance
```{r gom-pp-zoo, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-pp_zoo_test("GOM", "winter", "pp")
p2<-pp_zoo_test("GOM", "spring", "pp")
p3<-pp_zoo_test("GOM", "summer", "pp")
p4<-pp_zoo_test("GOM", "fall", "pp")

(p1 | p2)/(p3 | p4)
```

### GB
```{r GB-pp, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE,fig.width=11, fig.height=12} 
p1<-plot30less("GB", "winter", "pp")
p2<-plot30less("GB", "spring", "pp")
p3<-plot30less("GB", "summer", "pp")
p4<-plot30less("GB", "fall", "pp")

(p1 | p2)/(p3 | p4)
```

#### PP ~ Zoo abundance
```{r gb-pp-zoo, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-pp_zoo_test("GB", "winter", "pp")
p2<-pp_zoo_test("GB", "spring", "pp")
p3<-pp_zoo_test("GB", "summer", "pp")
p4<-pp_zoo_test("GB", "fall", "pp")

(p1 | p2)/(p3 | p4)
```


## CHL a Log Transformed {.tabset .tabset-fade}

### MAB
```{r mab-chl_log, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30less("MAB", "winter", "chl_log")
p2<-plot30less("MAB", "spring", "chl_log")
p3<-plot30less("MAB", "summer", "chl_log")
p4<-plot30less("MAB", "fall", "chl_log")

(p1 | p2)/(p3 | p4)
```

### GOM
```{r GOM-chl_log, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30less("GOM", "winter", "chl_log")
p2<-plot30less("GOM", "spring", "chl_log")
p3<-plot30less("GOM", "summer", "chl_log")
p4<-plot30less("GOM", "fall", "chl_log")

(p1 | p2)/(p3 | p4)
```

### GB
```{r GB-chl_log, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30less("GB", "winter", "chl_log")
p2<-plot30less("GB", "spring", "chl_log")
p3<-plot30less("GB", "summer", "chl_log")
p4<-plot30less("GB", "fall", "chl_log")

(p1 | p2)/(p3 | p4)
```


## PPD Log Transformed {.tabset .tabset-fade}

### MAB
```{r mab-pp_log, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30less("MAB", "winter", "pp_log")
p2<-plot30less("MAB", "spring", "pp_log")
p3<-plot30less("MAB", "summer", "pp_log")
p4<-plot30less("MAB", "fall", "pp_log")

(p1 | p2)/(p3 | p4)
```

### GOM
```{r GOM-pp_log, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30less("GOM", "winter", "pp_log")
p2<-plot30less("GOM", "spring", "pp_log")
p3<-plot30less("GOM", "summer", "pp_log")
p4<-plot30less("GOM", "fall", "pp_log")

(p1 | p2)/(p3 | p4)
```


### GB
```{r GB-pp_log, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE,fig.width=11, fig.height=12} 
p1<-plot30less("GB", "winter", "pp_log")
p2<-plot30less("GB", "spring", "pp_log")
p3<-plot30less("GB", "summer", "pp_log")
p4<-plot30less("GB", "fall", "pp_log")

(p1 | p2)/(p3 | p4)
```



## PAR {.tabset .tabset-fade}

### MAB
```{r mab-par, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30less("MAB", "winter", "par")
p2<-plot30less("MAB", "spring", "par")
p3<-plot30less("MAB", "summer", "par")
p4<-plot30less("MAB", "fall", "par")

(p1 | p2)/(p3 | p4)
```

### GOM
```{r GOM-par, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30less("GOM", "winter", "par")
p2<-plot30less("GOM", "spring", "par")
p3<-plot30less("GOM", "summer", "par")
p4<-plot30less("GOM", "fall", "par")

(p1 | p2)/(p3 | p4)
```


### GB
```{r GB-par, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE,fig.width=11, fig.height=12} 
p1<-plot30less("GB", "winter", "par")
p2<-plot30less("GB", "spring", "par")
p3<-plot30less("GB", "summer", "par")
p4<-plot30less("GB", "fall", "par")

(p1 | p2)/(p3 | p4)
```






## Bottom Temp {.tabset .tabset-fade}

### MAB
```{r mab-bt, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30less("MAB", "winter", "bt")
p2<-plot30less("MAB", "spring", "bt")
p3<-plot30less("MAB", "summer", "bt")
p4<-plot30less("MAB", "fall", "bt")

(p1 | p2)/(p3 | p4)
```

### GOM
```{r GOM-bt, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE, fig.width=11, fig.height=12} 
p1<-plot30less("GOM", "winter", "bt")
p2<-plot30less("GOM", "spring", "bt")
p3<-plot30less("GOM", "summer", "bt")
p4<-plot30less("GOM", "fall", "bt")

(p1 | p2)/(p3 | p4)
```


### GB
```{r GB-bt, echo = FALSE, eval=TRUE, warning = FALSE, message=FALSE,fig.width=11, fig.height=12} 
p1<-plot30less("GB", "winter", "bt")
p2<-plot30less("GB", "spring", "bt")
p3<-plot30less("GB", "summer", "bt")
p4<-plot30less("GB", "fall", "bt")

(p1 | p2)/(p3 | p4)
```







## Data summary


### SST
The data used to build plots comes from NOAA's physical science lab.  https://psl.noaa.gov/cgi-bin/db_search/DBListFiles.pl?did=132&tid=88124&vid=2423.


spatial range = 0.25 degree latitude x 0.25 degree longitude global grid (1440x720) 
temporaly range = 1982-2019. 

Table of seasonal ranges
1:90 <- Winter
91:181 <- Spring
182:273 <- Summer
274:365 <- Fall

Seasonal mean value calculated for each 1/4 degree raster. 

### Chla and pp 
Ask Kim H


### Bottom Temp 
[GLORYS](https://resources.marine.copernicus.eu/?option=com_csw&task=results?option=com_csw&view=details&product_id=GLOBAL_REANALYSIS_PHY_001_030) 
## Data analysis
Analysis run using `spatialwarnings`. 